\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-10 papers/paper1/ms.tex
adding some discussion of pre-conditioned algorithm

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{0,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& f_{n-1,j} + X_{n-1,j}\,z_{n-1} \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} U_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
    f_{n,j} &=& f_{n+1,j} + U_{n+1,j}\,b_{n+1} \nonumber\\
    b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} X_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

>>>>

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm t}$, where $t$ can be any arbitrarily large
real number.
This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{X}_{n-1,j}\,z_{n-1}\right] \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{X}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

\subsection{Simulation}
